---
title: "Multilevel Models using lmer"
author: "Joshua F. Wiley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multilevel-lmer-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how to use the `multilevelTools` package for
further diagnostics and testing of mixed effects (a.k.a., multilevel)
models using `lmer()` from the `lme4` package.

To get started, load the `lme4` package, which actually fits the
models, and the `multilevelTools` package. Although not required, we
load the `lmerTest` package to get approximate degrees of freedom for
use in calculating p-values for the fixed effects. Without `lmerTest`
p-values are based on asymptotic normality.
We'll also load an example dataset from the `JWileymisc` package,
`aces_daily`, using the `data()` function. The `aces_daily` dataset is
simulated data from a daily, ecological momentary assessment study of
191 participants who completed ratings of **a**ctivity, **c**oping,
**e**motions, and **s**tress (aces) up to three times per day for
twelve days. Only a subset of variables were simulated, but it is a
nice example of messy, real-world type data for mixed effects /
multilevel models.

```{r setup}

## load lme4, JWileymisc, and multilevelTools packages
## (i.e., "open the 'apps' ") 
library(lme4)
library(lmerTest)
library(extraoperators)
library(JWileymisc)
library(multilevelTools)

## load some sample data for examples
data(aces_daily, package = "JWileymisc")

```

## A Quick View of the Data

Let's start with a quick view of the **str**ucture of the data.

```{r}

## overall structure
str(aces_daily, nchar.max = 30)

```

For this example, we will just work with negative affect, `NegAff`, as
the outcome variable and stress, `STRESS`, as our predictor. Because
observations are repeated per person, we also need to use a variable
that indicates which observations belong to which person, `UserID`.
The code that follows shows how many unique people and observations we
have for each variable.

```{r}

## how many unique IDs (people) are there?
length(unique(aces_daily$UserID))

## how many not missing observations of negative affect are there?
sum(!is.na(aces_daily$NegAff))

## how many not missing observations of stress are there?
sum(!is.na(aces_daily$STRESS))

```

Finally, we might explore each variable briefly, before jumping into
analyzing them and examining diagnostics and effects of our models.
Using the `summary()` function on each variable gives us some basic
descriptive statistics in the form of a five number summary (minimum,
first quartile, median (second quartile), third quartile, and
maximum) as well as the arithmetic mean. The NA's tell us how many
observations are missing negative affect. We can see that both
variables are quite skewed as their minimum, first quartile and
medians are all close together and far away from the maximum.

```{r}

summary(aces_daily$NegAff)

summary(aces_daily$STRESS)

```

Because these are repeated measures data, another useful descriptive
statistic is the intraclass correlation coefficient or ICC. The ICC is
a measure of the proportion of variance that is between people versus
the total variance (i.e., variance between people and variance within
persons). `multilevelTools` provides a function, `iccMixed()` to
estimate ICCs based off of mixed effects / multilevel models. The
following code does this for negative affect and stress, first naming
all the arguments and then using shorter unnamed approach, that is
identical results, but easier to type. The relevant output is the ICC
for the row named `UserID`. An ICC of 1 indicates that 100% of all
variance exists between people, which would mean that 0% of variance
exists within person, indicating that people have identical scores
every time they are assessed. Conversely an ICC of 0 would indicate
that everyone's average was identical and 100% of the variance exists
within person. For negative affect and stress, we can see the ICCs
fall between 0 and 1, indicating that some variance is between people
(i.e., individuals have different average levels of negative affect
and stress) but also that some variance is within person, meaning that
people's negative affect and stress fluctuate or vary within a person
across the day and the 12-days of the study.

```{r}

iccMixed(
  dv = "NegAff",
  id = "UserID",
  data = aces_daily)

iccMixed("STRESS", "UserID", aces_daily)

``` 

Finally, we might want to examine the distribution of the variables
visually. Visual exploration is a great way to identify the
distribution of variables, extreme values, and other potential issues
that can be difficult to identify numerically, such as bimodal
distributions. For multilevel data, it is helpful to examine between
and within person aspects of a variable separately. `multilevelTools`
makes this easy using the `meanDecompose()` function. This is
important as, for example, if on 11 of 12 days, someone has a negative
affect score of 5, and then one day a score of 1, the score of 1 may
be an extreme value, for that person even though it is common for the
rest of the participants in the study. `meanDecompose()` returns a
list with `X` values at different levels, here by ID and the
residuals, which in this case are within person effects.

We make plots of the distributions using `testDistribution()`, which
defaults to testing against a normal distribution, which is a common
default and in our case appropriate for linear mixed effects /
multilevel models. The graphs show a density plot in black lines, a
normal distribution in dashed blue lines, a rug plot showing where
individual observations fall, and the x-axis is a five number summary
(minimum, first quartile, median, third quartile, maximum). The bottom
plot is a Quantile-Quantile plot rotated to be horizontal instead of
diagonal. Black dots / lines indicate extreme values, based on the
expected theoretical distribution, here normal (gaussian), and
specified percentile, `.001`. For more details, see 
`help("testDistribution")`.

The between person results for negative affect show about five people
with relative extreme higher scores. Considering negative affect
ranges from 1 to 5, they are not impossible scores, but they are
extreme relative to the rest of this particular sample. At the within
person level, there are thousands of observations and there are many
extreme scores. Based on these results, we may suspect that when we
get into our mixed effects / multilevel model, there will be more
extreme residual scores to be dealt with (which roughly correspond to
within person results) than extreme random effects (which roughly
correspond to between person results).

```{r} 

tmp <- meanDecompose(NegAff ~ UserID, data = aces_daily)
str(tmp, nchar.max = 30)

plot(testDistribution(tmp[["NegAff by UserID"]]$X,
                      extremevalues = "theoretical", ev.perc = .001),
     varlab = "Between Person Negative Affect")

plot(testDistribution(tmp[["NegAff by residual"]]$X,
                      extremevalues = "theoretical", ev.perc = .001),
     varlab = "Within Person Negative Affect")

``` 

The same exercise for stress shows a better between and within person
distribution. Certainly there is some skew and its not a perfect
normal distribution. However, neither of these are required for
predictors in mixed effects models. The extreme values are not that
extreme or that far away from the rest of the sample, so we might feel
comfortable proceeding with the data as is.

```{r}

tmp <- meanDecompose(STRESS ~ UserID, data = aces_daily)

plot(testDistribution(tmp[["STRESS by UserID"]]$X,
                      extremevalues = "theoretical", ev.perc = .001),
     varlab = "Between Person STRESS")

plot(testDistribution(tmp[["STRESS by residual"]]$X,
                      extremevalues = "theoretical", ev.perc = .001),
     varlab = "Within Person STRESS")


``` 

# Mixed Effects (Multilevel) Model

With a rough sense of our variables, we proceed to fitting a mixed
effects model using the `lmer()` function from `lme4`. Here we use
negative affect as the outcome predicted by stress. Both the intercept
and linear slope of stress on negative affect are included as fixed
and random effects. The random effects are allowed to be correlated,
so that, for example, people who have higher negative affect under low
stress may also have a flatter slope as thye may have less room to
increase in negative affect at higher stress levels.

At the time of writing, this model using the default optimizer and
control criteria fails to converge (albeit the gradient is very
close). These convergence warnings can be spurious but also may
suggest either a too complex model or need to refine some other aspect
of the model, such as scaling predictors or changing the optimizer or
arguments to the optimizer.

```{r, warning = TRUE, purl = FALSE}

m <- lmer(NegAff ~ STRESS + (1 + STRESS | UserID),
          data = aces_daily)

```

Here we use the `lmerControl()` function to specify a different
optimization algorithm using the Nelder-Mead Method 
(https://en.wikipedia.org/wiki/Nelder%E2%80%93Mead_method). We also
tighten the tolerance values. This is then passed to `lmer()` by
setting the argument, `control = strictControl`. This converges
without warning. 

```{r}

strictControl <- lmerControl(optCtrl = list(
   algorithm = "NLOPT_LN_NELDERMEAD",
   xtol_abs = 1e-12,
   ftol_abs = 1e-12))

m <- lmer(NegAff ~ STRESS + (1 + STRESS | UserID),
          data = aces_daily, control = strictControl)

```

Now we move on to examine model diagnostics. For linear mixed effects
/ multilevel models, the residuals should follow a normal
distribution, the random effects should follow a multivariate normal
distribution, and the residual variance should be homogenous (the
same) as a single residual variance is estimated and used for the
whole model. The `modelDiagnostics()` function in `multilevelTools`
helps to evaluate these assumptions graphically.

The first plot shows (top left) the distribution of residuals. It is
somewhat too narrow for a normal distribution (i.e., leptokurtic) and
consequently many of the residuals on both tails are considered
extreme values. However, the distribution is fairly symmetrical and
for the residuals we have over 6,000 observations which will tend to
make results relatively robust to violations.

The second plot (top right) shows the fitted (predicted) values for
each observation against the residuals. The colour of the blocks
indicates how many observations fall at a particular point. The solid
blue line is a loess smooth line. Hopefully this line is about flat
and stays consistently at residuals of 0 regardless of the predicted
value, indicating no systematic bias. Finally the dashed blue lines
indicate the 10th and 90th percentile (estimated from a quantile
regression model) of the residuals across predicted values. If the
residual variance is homogenous across the spread of predicted values,
we would expect these dashed lines to be flat and parallel to each
other. That is approximately true for all predicted values above about
1.7. For low predicted values, the residuals are 0 or higher and the
spread of the dashed lines is narrowed. This is because people cannot
have negative affect scores below 1 and many people reported negative
affect around 1. Therefore for predicted negative affect scores of 1,
the residuals tend to be about 0. This pattern is not ideal, but also
not indicative of a terrible violation of the assumption. Two possible
approaches to relaxing the assumption given data like these would be
to use a censored model, that assumes true negative affect scores may
be lower but are censored at 1 due to limitations in the
measurement. Another would be to use a location and scale mixed
effects model that models not only the mean (location) but also the
scale (variance) as a function of some variables, which would allow
the residual variance to differ. These are beyond the scope of this
document, however.

The last three plots show the univariate distribution of the intercept
and stress slope by `UserID`, the random intercept and slope and a
test of whether the random intercept and slope are multivariate normal
The multivariate normality test, based on the mahalonbis distances,
suggests that there are a few, relatively extreme people. We might
consider dropping these individuals from the analysis to examine
whether results are sensitive to these extreme cases.

```{r, fig.width = 7, fig.height = 10}

md <- modelDiagnostics(m, ev.perc = .001)

plot(md, ask = FALSE, ncol = 2, nrow = 3)

```

In order to drop people who were extreme on the multivariate normality
test, we can access a data table of the extreme values from the
`modelDiagnostics` object. We subset it to only include the
multivariate effects and use the `head()` function to show the first
few rows. The extreme value table shows the scores on the outcome, the
`UserID`, the index (which row) in the dataset, and the effect type,
here all multivariate since we subset for that. We can use the 
`unique()` function to identify the IDs, which shows it is three IDs.

```{r}

mvextreme <- subset(md$extremeValues,
  EffectType == "Multivariate Random Effect UserID")

head(mvextreme)

unique(mvextreme$UserID)

```

We can update the existing model using the `update()` function and in
the data, just subset to exclude those three extreme IDs. We re-run
the diagnostics and plot again revealing one new multivariate extreme
value. 

```{r, fig.width = 7, fig.height = 10}

m2 <- update(m, data = subset(aces_daily,
  UserID %!in% unique(mvextreme$UserID)))

md2 <- modelDiagnostics(m2, ev.perc = .001)

plot(md2, ask = FALSE, ncol = 2, nrow = 3)

mvextreme2 <- subset(md2$extremeValues,
  EffectType == "Multivariate Random Effect UserID")

unique(mvextreme2$UserID)

```

Again removing this further extreme ID and plotting diagnostics
suggests that now the random effects are fairly "clean".


```{r, fig.width = 7, fig.height = 10}

m3 <- update(m, data = subset(aces_daily,
  UserID %!in% c(unique(mvextreme$UserID), unique(mvextreme2$UserID))))

md3 <- modelDiagnostics(m3, ev.perc = .001)

plot(md3, ask = FALSE, ncol = 2, nrow = 3)

```

Now that we have a model whose diagnostics we are reasonably happy
with, we can examine the results. The `modelPerformance()` function
from the `multilevelTools` package gives some fit indices for the
overall model, including measures of the variance accounted for by the
fixed effects (marginal R2) and from the fixed and random effects
combined (conditional R2). We also get information criterion (AIC,
BIC), although note that with a REML estimator, the log likelihood and
thus information criterion are not comparable to if the ML estimator
was used.

```{r}

modelPerformance(m3)

```

To see the results of individual variables, we can use the `summary()`
function to get the default model summary. Note that this summary
differs slightly from that produced by `lme4` as it is overridden by
`lmerTest` which adds degrees of freedom and p-values.

```{r}

summary(m3)

``` 

This default summary gives quite a bit of key information. However, it
does not provide some of the results that often are desired for
scientific publication. Confidence intervals are commonly reported and
t values often are not reported in preference for p-values. In
addition, it is increasingly common to ask for effect sizes.
The `modelTest()` function in `multilevelTools` provides further
tests, including tests of the combined fixed + random effect for each
variable and effect sizes based off the independent change in marginal
and conditional R2, used to calculate a sort of cohen's F2.
All of the results are available in a series of tables for any
programattic use. However, for individual use or reporting, caling
`APAStyler()` will produce a nicely formatted output for humans.
Confidence intervals are added in brackets and the effect sizes at the
bottom are listed for stress considering fixed + random effects
together.


```{r}

mt3 <- modelTest(m3)

names(mt3) ## list of all tables available

APAStyler(mt3)

``` 

The output of `APAStyler()` can easily be copied and pasted into Excel
or Word for formatting and publication. If the format is not as
desired, some changes are fairly easy to make automatically.
For example the following code uses 3 decimal points, lists exact
p-values, and uses a semi colon instead of a comma for confidence
intervals.

```{r}

APAStyler(mt3,
          format = list(
            FixedEffects = "%s, %s (%s; %s)",
            RandomEffects = c("%s", "%s (%s, %s)"),
            EffectSizes = "%s, %s; %s"),
          digits = 3,
          pcontrol = list(digits = 3, stars = FALSE,
                          includeP = TRUE, includeSign = TRUE,
                          dropLeadingZero = TRUE))

```

Finally, `APAStyler()` can be used with multiple models to create a
convenient comparison. For example, although we removed several
extreme values, we might want to compare the results in the full data
to the dataset with extreme values removed to evaluate whether
critical coefficients or effect sizes changed and if so how much.
Some quantities, like the log likelihood, AIC and BIC are not
comparable as the sample size changed. However, effect size estimates
like the model marginal and conditional R2 can be reasonably compared
as can the fixed effect coefficients and the effect sizes for
particular predictors.

To create the output, we pass a list of `modelTest` objects and we can
add names so that they are more nicely named in the output. The
results show that the intercept, the predicted negative affect when
stress is zero is very slightly higher in the original than in the
model with outliers removed. The fixed effect coefficient for stress
is identical, but the confidence interval is very slightly wider after
removing outliers. Variability in the random intercept, slope, and
residual variance, sigma, are all slightly reduced. All in all, we
could conclude that the overall pattern of results and any conclusions
that might be drawn from them would not functionally change in the
model with all cases included versus after removing the extreme
values, in this case. This effectively serves as a "sensitivity
analysis" evaluating how sensitive the results of the model are to the
inclusion / exclusion of extreme values. In this case, not very
sensitive, which may be encouraging. In cases where there are large
differences in the results, careful thought may be needed around which
model is more likely to be "true" and what the implications of the
differences are for interpretting and utilizing the results.

```{r}
## run modelTest() on the original model, m
mt <- modelTest(m)

APAStyler(list(Original = mt, `Outliers Removed` = mt3))

``` 
